\section{Introduction}
\label{sec:intro}

Diet-related chronic disease is one of the largest modifiable drivers of population health, and smartphone cameras have made it trivial to record what we eat.
What has proven stubbornly hard is converting a single photograph of a plate into actionable feedback:
how balanced is this meal, and does it resemble the kind of plate nutritional guidelines recommend?
A growing body of computer-vision work targets pieces of this problem---food classification~\cite{bossard2014food101}, detection~\cite{matsuda2012uecfood}, and segmentation~\cite{wu2021foodseg103}---but most published pipelines stop at \emph{identification} and leave the leap from ``broccoli, rice, chicken'' to ``was this a good meal?'' to the user.

\textbf{SmartPlate.}
We tackle the full pipeline from RGB image to dietary feedback in a deliberately lightweight way.
Given a single plate photo, we (i) predict a pixel-level segmentation across $103$ fine-grained food categories using a SegFormer-B0~\cite{xie2021segformer} fine-tuned on FoodSeg103~\cite{wu2021foodseg103}, (ii) aggregate the per-pixel predictions into $7$ food groups drawn from the Harvard Healthy Eating Plate~\cite{harvardplate}, and (iii) score each plate on a $0$--$100$ scale using a transparent, rule-based rubric that rewards meeting the Harvard target proportions and penalizes refined grains, red or processed meat, and sugary/fatty items.
The same pixel-area proportions drive a coarse nutritional breakdown through a per-group USDA~\cite{usdafdc} macro lookup.

Our framing deliberately favors an \emph{interpretable} scoring module over a learned regressor: the score comes from a short formula over food-group proportions, so the reason a plate receives the score it does is trivially auditable.
This keeps the learned component of the system narrowly scoped to the part of the problem where learning adds real value---\emph{where on the image is the broccoli?}---and leaves dietary judgement to a published rubric.

\textbf{Contributions.}
\begin{enumerate}
    \item An end-to-end pipeline from a raw plate image to a Harvard-aligned healthy-plate score plus a coarse nutritional breakdown, all runnable on a single Apple Silicon laptop.
    \item A fine-tuned SegFormer-B0 food-segmentation model with quantitative evaluation at both $103$-class and $7$-group granularity on the FoodSeg103 validation split.
    \item A ResNet-50~\cite{he2016resnet} multi-label classification baseline trained on the same data, giving a controlled comparison of what pixel-level outputs buy us over image-level outputs for this task.
    \item An interpretable, closed-form scoring function, plus a hand-built mapping from the $103$ FoodSeg103 classes to the $7$ Harvard food groups that we release alongside the code.
\end{enumerate}
